---
title: "Analysis of Cyclistic's Casual Riders vs Annual Members"
author: "Muhaimin Ahmed"
date: "August 28, 2024"
output: 
  github_document: default
---

# Introduction

This is a CAPSTONE case study I did as part of my Google Data Analytics Certification. In this case study, I work for a fictional company, Cyclistic and have to answer business questions from my manager Lily Moreno.

## About Cyclistic

Cyclistic is a bike-share program that features more than 5,800 bicycles and 600 docking stations. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, I want to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, I will design a new marketing strategy to convert casual riders into annual members.

## Guiding Questions

The questions guiding this analysis are the following:

1.  How do annual members and casual riders use Cyclistic bikes differently?
2.  Why would casual riders buy Cyclistic annual memberships?
3.  How can Cyclistic use digital media to influence casual riders to become members?

## Location of the Data

The data used for this analysis is located [here](https://divvy-tripdata.s3.amazonaws.com/index.html). From the available data the 12 most recent months have been chosen for analysis (August 2023 to August 2024).

## Business Task

Identify potential opportunities for growth and recommendations for the Cyclistic marketing strategy to convert more casual riders to become members.

# Preparing the Data

## Libraries

These are the libraries that are used to complete this case study.

```{r adding libraries}
library(tidyverse)
library(dplyr)
library(readr)
library(lubridate)
library(ggplot2)
```

## Merging the Data

12 months of data are merged onto one data frame named cyclistic_df.

```{r load data and merge onto one df}
# temp <- list.files("./Cyclistic_Data", full.names = TRUE, pattern = "\\.csv$")
# 
# cyclistic_df <- readr::read_csv(temp, id = "Bike_Rides_2306_to_2407")
# 
# write.csv(cyclistic_df, "./Cyclistic_Data/cyclistic_merged_data.csv")

cyclistic_df <- readr::read_csv("./Cyclistic_Data/cyclistic_merged_data.csv")
```

Let's focus on the following columns to start the analysis:

-   ride_id: Ride unique ID.
-   member_casual: Cyclistic member or casual rider.
-   started_at and ended_at: Ride start and end times.
-   rideable_type: Electric, classic or docked bike.

```{r selecting start and stop times}
cyclistic_df_v2 <- select(cyclistic_df, ride_id, member_casual, started_at, ended_at, rideable_type)
head(cyclistic_df_v2)
```

## Checking for Missing Values

```{r}
sum(is.na(cyclistic_df_v2))
```

The data is complete for the columns of interest.

# Summarizing the Data

Using the selected fields the ride length can be calculated and the day of the week can also be determined and store in cyclistic_df_v2. I am also adding a month year column and sorting it chronologically for later use.

```{r add ride length and day of the week}
cyclistic_df_v3 <- mutate(cyclistic_df_v2, 
                          ride_lengths = abs(seconds_to_period(ended_at-started_at)), 
                          day_of_week = weekdays(started_at), 
                          month_year = format(started_at, "%b-%y")
                          ) %>%
  arrange(started_at) %>%  # Arrange the data by the original date-time column
  mutate(
    month_year = factor(month_year, levels = unique(month_year), ordered = TRUE)
  )
head(cyclistic_df_v3)
```

# Analyzing the data

```{r max ride length}
max(cyclistic_df_v3$ride_lengths)
```

```{r mean ride length}
mean(cyclistic_df_v3$ride_lengths)
```

```{r mode day of week and  rideable type}
mode_char <- function(x) unique(x)[which.max(tabulate(match(x, unique(x))))]
mode_char(cyclistic_df_v3$day_of_week)
mode_char(cyclistic_df_v3$rideable_type)
```

```{r average ride length for member and casual riders}
cyclistic_df_v3 %>%
  group_by(member_casual) %>%
  summarize(avg_ride_length = mean(ride_lengths, na.rm = TRUE))
```

We can conclude the following so far:

-   The longer ride was about an hour.
-   Average ride taking approximately 30 minutes.
-   Saturday is the most popular day for all users to ride bikes.
-   Electric bikes are more used than classic bikes.
-   On average there is no significant difference between casual and member users for ride length.

## Further Summarizing the Data

```{r distribution of bike types used by member and casual riders}

bike_type_by_member <- cyclistic_df_v3 %>%
  group_by(member_casual, rideable_type) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = rideable_type, values_from = count, values_fill = 0) %>%
  gather(key = "bike_type", value = "count", -member_casual)
head(bike_type_by_member)
```

# Visualization

```{r}
ggplot(bike_type_by_member, aes(x = member_casual, y = count, fill = bike_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Distribution of Bike Types by Member Type",
       x = "Member Type",
       y = "Count",
       fill = "Bike Type") + 
  theme(plot.title = element_text(hjust = 0.5))
```

We can see that there are already much more member users than there are casual users. However, both casual and member users use the classic and electric bikes relatively equally. This may mean that it is based on availability over preference. However more data is required to derive further conclusions from this graph.

```{r average ride length and number of rides for users by day of week}
summarized_cyclistic_data <- cyclistic_df_v3 %>%
  group_by(day_of_week, member_casual) %>%
  summarize(avg_ride_length = mean(ride_lengths), 
            num_rides = n())
```

```{r}
summarized_cyclistic_data$day_of_week <- factor(summarized_cyclistic_data$day_of_week, c("Monday", "Tuesday", "Wednesday", "Thursday","Friday","Saturday", "Sunday"))
```

```{r}
head(summarized_cyclistic_data)
```

```{r average ride length for users by day of week}
ggplot(data = summarized_cyclistic_data, aes(x = day_of_week, y = avg_ride_length, fill = member_casual)) +
  geom_bar(position = "dodge", stat = "identity") +
  labs(title = "Average Ride Length for Users by Day of the Week",
       x = "Day of the week",
       y = "Average Ride Length",
       fill = "User Type") + 
  theme(plot.title = element_text(hjust = 0.5))
```

As discussed before there is no significant difference in ride length between casual and member users. This graph clarifies this further by showing the it fluctuates very little throught the week.

```{r number of rides for users by day of week}
ggplot(data = summarized_cyclistic_data, aes(x = day_of_week, y = num_rides, fill = member_casual)) +
  geom_bar(position = "dodge", stat = "identity") +
  labs(title = "Number of Rides for Users by Day of the Week",
       x = "Day of the week",
       y = "Number of Rides",
       fill = "User Type") + 
  theme(plot.title = element_text(hjust = 0.5))
```

This graph shows that more member users ride during the week than in the weekend and there are more casual rider trips during the weekend. This means that member users use the service to commute to work while casual users use it leisurely on the weekend. Also as it was seen in the first graph, even though there are more casual riders in the weekend it is still less than the number of member users.

```{r monthly summary}
monthly_cyclistic_data <- cyclistic_df_v3 %>%
  group_by(month_year, member_casual) %>%
  summarize(avg_ride_length = mean(ride_lengths), 
            num_rides = n())

head(monthly_cyclistic_data)
```

I was also curious to see the number of rides throughout a year. And unsurprisingly there are more rides in the warmer months than there are in the colder months. Casual riders drop off much more than the member users. Once again it comes back to the difference in the reason why the two types of users use the service. Casual users use it for leisure so they can afford not biking as much in the colder months, however more member users may use this service for commute to work which can be why their numbers don't drop as low.

```{r number of rides for users by month}
ggplot(data = monthly_cyclistic_data, aes(x = month_year, y = num_rides, fill = member_casual)) +
  geom_bar(position = "dodge", stat = "identity") +
  labs(title = "Number of Rides for Users by Month",
       x = "Month and Year",
       y = "Number of Rides",
       fill = "User Type") + 
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_text(angle = 45, size = 8))
```

# Conclusion

We can now use our analysis to answer the aforementioned guiding questions.

## Difference between annual and casual riders.

According to the data, more annual members ride during the week than the weekend. This may mean that annual members use Cyclistic to commute to work during the week which is why they are less likely to use it during the weekend. On the other hand, more casual riders use Cyclistic during the weekend. This hints at casual riders using Cyclistic for leisurely purposes during their time off in the weekend.

## Reasons to switch to an annual membership for the casual rider.

Casual riders may consider purchasing Cyclistic annual memberships for several reasons:

1.  Cost Savings: If a casual rider finds themselves using Cyclistic more frequently over time, they might realize that the cost of individual rides adds up. An annual membership could be more cost-effective if they plan to use the service regularly, even if primarily on weekends.

2.  Flexibility and Convenience: An annual membership offers the flexibility to ride anytime without worrying about the cost of each trip. For riders who value convenience and spontaneity, an annual membership removes the hassle of repeatedly purchasing single rides.

3.  Increased Usage Over Time: Casual riders who initially use Cyclistic for leisurely weekend rides might develop a habit or enjoy the experience enough to start using it more frequently, perhaps even incorporating it into their weekly routine. An annual membership could be a logical step as their usage increases.

4.  Commitment to a Healthier Lifestyle: Some casual riders might see an annual membership as a commitment to a healthier, more active lifestyle. Regular biking, even if only on weekends, can be part of a fitness routine or a way to explore the city.

5.  Special Promotions or Incentives: Cyclistic might offer promotions, discounts, or added benefits that make an annual membership more appealing to casual riders. For example, discounts on annual memberships during certain times of the year could encourage casual riders to commit.

6.  Supporting a Green Initiative: Riders who are environmentally conscious might choose to buy an annual membership as a way of supporting sustainable transportation options, even if their usage is primarily for leisure.

7.  Peer Influence: If a rider’s friends or family members have annual memberships, they might be encouraged to get one themselves to join group rides or participate in shared activities.

These factors can make the annual membership appealing to casual riders, even if their primary usage remains centered around the weekend.

## Recomendation to incentivize casual to member rider conversion.

Cyclistic can leverage digital media to influence casual riders to become members through several strategic approaches:

1.  Targeted Advertising:

-   Social Media Ads: Use platforms like Instagram, Facebook, and Twitter to run targeted ads that highlight the benefits of an annual membership. These ads can be customized based on user behavior, such as those who have previously rented bikes or visited the Cyclistic website.
-   Google Ads: Implement search and display ads targeting keywords like "bike rental," "weekend activities," or "leisure biking." These ads can direct users to landing pages that explain the advantages of a membership.

2.  Email Marketing:

-   Personalized Offers: Send personalized emails to casual riders with exclusive discounts or promotions for upgrading to an annual membership. Emails can include usage statistics showing how much they could save with a membership.
-   Usage-Based Recommendations: Based on data, email casual riders after a certain number of trips, highlighting how close they are to the cost of an annual membership and offering a discount to close the gap.

3.  Engaging Content:

-   Blogs and Articles: Create blog posts or articles on the Cyclistic website discussing topics like “Top 10 Weekend Biking Trails,” subtly promoting how an annual membership can enhance these experiences.
-   Social Media Content: Share success stories or testimonials from casual riders who upgraded to an annual membership, emphasizing the benefits they’ve experienced, like cost savings or improved health.
-   Video Content: Produce short, engaging videos for platforms like YouTube or TikTok, showcasing real-life scenarios where casual riders benefit from an annual membership, such as commuting to work, weekend outings, or spontaneous rides.

4.  Referral Programs:

-   Incentivized Referrals: Launch a referral program where current members can invite casual riders to join with a discount on their first annual membership. Reward both the referrer and the new member with additional ride credits or discounts.

5.  Social Proof and Influencer Partnerships:

-   Testimonials and Reviews: Encourage satisfied members to share their experiences on review sites, social media, or through testimonials featured on the Cyclistic website.
-   Influencer Collaborations: Partner with local influencers who align with the brand’s values (e.g., fitness enthusiasts, eco-friendly advocates) to promote the annual membership through their platforms, providing them with free memberships or other incentives.

6.  Interactive Campaigns and Challenges:

-   Biking Challenges: Create digital challenges that encourage casual riders to track their rides through an app, offering rewards for milestones that could be best utilized with an annual membership.
-   Gamification: Incorporate gamification elements in an app where users earn points or badges for rides taken, with rewards or discounts on annual memberships once certain thresholds are met.

7.  Retargeting Strategies:

-   Website Retargeting: Use cookies to track casual riders who have visited the Cyclistic website or app but haven’t purchased a membership. Serve them retargeted ads on other websites and social media platforms, emphasizing the benefits of an annual membership.
-   Abandoned Cart Emails: If Cyclistic offers online membership purchases, send automated emails to users who begin the membership purchase process but do not complete it, reminding them of the benefits and offering a limited-time discount.

8.  Exclusive Member Content:

-   Webinars and Virtual Events: Host webinars or virtual events discussing the best biking routes, maintenance tips, or health benefits of biking, offering special content or early access to members. Promote these events to casual riders to entice them to join for exclusive access.

By using these digital media strategies, Cyclistic can create a compelling narrative and offer tangible incentives for casual riders to transition to annual memberships, ultimately fostering greater brand loyalty and increased revenue.

